Fair Queuing Aware Congestion Control

### 2022-06-16 picoquic/tex/paper.tex
Proofread again

\begin{abstract}
Fair queuing is becoming increasingly prevalent in the internet and has been shown to improve performance in many circumstances. 
Performance could be improved even more if endpoints could detect the presence of fair queuing on a certain path and adjust their congestion control accordingly. 
If fair queuing is detected, the congestion control would not have to take cross traffic into account, which allows for more flexibility. 
In this paper, we develop the first algorithm that continuously checks if fair queuing is present on a path. 
When fair queuing is detected, a different congestion control is chosen, which results in reduced latency. 
Unlike an algorithm proposed in a previous paper of us, the approach presented here does not only detect the presence of fair queuing once at flow startup but it does so continuously. 
% We show that by adding fair queuing detection to an existing congestion control algorithm, queuing delay can be reduced by more than two thirds. 
\end{abstract}

@@

While fair queuing was proposed decades ago, it only gained popularity in the last couple of years because of implementations in the Linux kernel \cite{dumazet_pkt_sched_2013,hoeiland-joergensen_flow_2018} and also in Apple's macOS. 
Applications can benefit from increasing deployment of fair queuing: It makes sure that not the most aggressive one wins. 
It would be even better if applications could know if the connection they're sending on is managed by fair queuing. 
Then they could be sure that the can use their preferred congestion control mechanism, while not bothering or being bothered by other network flows. 
We proposed the first such approach in previous work \cite{bachl_detecting_2021} but our previous approach had some shortcomings upon which we improve in this paper.

In our previous work we proposed a technique which determines the presence of fair queuing at flow startup, which worked as follows: 
If fair queuing is successfully detected at flow startup, a congestion control was used, which aimed to keep queuing delay low (delay-based congestion control). 
While this delay-based congestion control achieved high throughput and low delay, it was vulnerable to be outcompeted by other network flows sending more aggressively, such as \cite{cardwell_bbr_2016,dong_pcc_2015,ha_cubic_2008}, 
similar to the Vegas congestion control algorithm \cite{brakmo_tcp_1995}.
This means that our delay-based congestion control performed well but only when it wouldn't have to compete with other flows. Thus is only used if fair queuing is detected. 
If the absence of fair queuing is detected, our previous approach uses a more aggressive congestion control (specifically PCC \cite{dong_pcc_2015}), which can compete better with other network flows,
but doesn't keep delay as low as our delay-based congestion control.

@@

The core concept of our approach, which we call \textbf{\textit{Tonopah}}, is that a network flow is separated in \textbf{two subflows}. 
\textbf{Alternatingly}, one flow sends more (\textbf{dominant subflow}), while the other one sends less (\textbf{non-dominant subflow}). 
After a certain time interval the dominant subflow becomes the non-dominant one and vice versa. 
When the total sending rate reaches the link capacity and fair queuing is present, 
\textbf{fair queuing} is going to \textbf{limit} the throughput of the \textbf{dominant subflow} (\autoref{fig:illustration_fq}). 
This can be measured by observing the \textit{delivery rate}
\footnote{The delivery rate (a concept popularized by BBR \cite{cardwell_bbr_2016}) can be measured by observing the rate at which acknowledgement packets are sent back to the sender by the receiver.
If the sender sends more than the link capacity, the delivery rate is never going to exceed the link capacity since the receiver cannot receive packets faster than the link capacity allows. 
}
of both subflows. If there is no fair queuing, the subflow which sends more also achieves a higher throughput (\autoref{fig:illustration_no_fq}), which can be considered unfair. 
In this case, Tonopah doesn't behave different from a network flow which doesn't have fair queuing detection.

@@

The dominant subflow receives $\frac{6}{10}$ of the total allowed sending rate, while the non-dominant one receives $\frac{4}{10}$. 
The dominant subflow and the non-dominant subflow alternate every $\max{(50, \min{(100, \textit{rtt})})}$ milliseconds. 
This time period is also called an \textbf{interval}. To determine the presence of fair queuing, the data of the last four intervals are evaluated. 
We look at more than one interval because this reduces the effect of measurement noise. We found four intervals to be a suitable number.

@@

Another safeguard against measurement noise is that we only take a decision regarding fair queuing if the dominant flow actually really sent more data than the non-dominant one. 
Specifically, if the dominant flow sends less than 57.5\%, we don't take a decision because it means that probably we didn't have enough data to send. 
This case only occurs when the sending rate is very low ($<$ 10\,Mbit/s).

@@

Tonopah was implemented on top of Picquic in C. It was added on top of the NewReno congestion control algorithm but it could easily be added on top of other 
congestion control algorithms as well. By basing our implementation on Picoquic, Tonopah also supports Explicit Congestion Notification.

@@

In the following evaluation we always run each network flow for 90\,s. Tonopah measures whether there is fair queuing or not continuously. 
If in the evaluation scenario there is fair queuing and Tonopah detects fair queuing for 80\,s and fails to detect it for 10\,s, we 
state the accuracy as $\frac{80}{90} = 89\%$. 
\begin{table}
\begin{tabularx}{\columnwidth}{| l | X | X | X |}
\hline
& 10 Mbit/s & 50 Mbit/s & 100 Mbit/s \\ \hline
10ms & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\%\\ \hline
50ms & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% \\ \hline
100ms & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% \\ \hline
\end{tabularx}
\caption{Detection accuracy in case there's \textbf{no fair queuing}.  The overall median accuracy is 100\%, the first quartile is 100\% and the third quartile is 100\%.}
\label{table:no_fq}
\end{table}

@@

>>>>

\begin{abstract}
Fair queuing is becoming increasingly prevalent in the internet and has been shown to improve performance in many circumstances. 
Performance could be improved even more if endpoints could detect the presence of fair queuing on a certain path and adjust their congestion control accordingly. 
If fair queuing is detected, the congestion control would not have to take cross traffic into account, which allows for more flexibility. 
In this paper, we develop the first algorithm that continuously checks if fair queuing is present on a path. 
When fair queuing is detected, a different congestion control can be chosen, which can result in reduced latency. 
Unlike an algorithm proposed in a previous paper of us, the approach presented here does not only detect the presence of fair queuing once at flow startup but it does so continuously. 
% We show that by adding fair queuing detection to an existing congestion control algorithm, queuing delay can be reduced by more than two thirds. 
\end{abstract}

@@

While fair queuing was proposed decades ago, it only gained popularity in the last couple of years because of implementations in the Linux kernel \cite{dumazet_pkt_sched_2013,hoeiland-joergensen_flow_2018} and also in Apple's macOS. 
Applications can benefit from increasing deployment of fair queuing: It makes sure that not the most aggressive one wins. 
It would be even better if applications could know if the connection they're sending on is managed by fair queuing. 
Then they could be sure that they can use their preferred congestion control mechanism, while not bothering or being bothered by other network flows. 
We proposed the first such approach in previous work \cite{bachl_detecting_2021} but our previous approach had some shortcomings upon which we improve in this paper.

In our previous work we proposed a technique which determines the presence of fair queuing at flow startup, which worked as follows: 
If fair queuing is successfully detected at flow startup, a congestion control was used, which aimed to keep queuing delay low (delay-based congestion control). 
While this delay-based congestion control achieved high throughput and low delay, it was vulnerable to be outcompeted by other network flows sending more aggressively, such as \cite{cardwell_bbr_2016,dong_pcc_2015,ha_cubic_2008}, 
similar to the Vegas congestion control algorithm \cite{brakmo_tcp_1995}.
This means that our delay-based congestion control performed well but only when it wouldn't have to compete with other flows. Thus it is only used if fair queuing is detected. 
If the absence of fair queuing is detected, our previous approach uses a more aggressive congestion control (specifically PCC \cite{dong_pcc_2015}), which can compete better with other network flows,
but doesn't keep delay as low as our delay-based congestion control.

@@

The core concept of our approach, which we call \textbf{\textit{Tonopah}}, is that a network flow is separated into \textbf{two subflows}. 
\textbf{Alternatingly}, one flow sends more (\textbf{dominant subflow}), while the other one sends less (\textbf{non-dominant subflow}). 
After a certain time interval the dominant subflow becomes the non-dominant one and vice versa. 
When the total sending rate reaches the link capacity and fair queuing is present, 
\textbf{fair queuing} is going to \textbf{limit} the throughput of the \textbf{dominant subflow} (\autoref{fig:illustration_fq}). 
This can be measured by observing the \textit{delivery rate}
\footnote{The delivery rate (a concept popularized by BBR \cite{cardwell_bbr_2016}) can be measured by observing the rate at which acknowledgement packets are sent back to the sender by the receiver.
If the sender sends more than the link capacity, the delivery rate is never going to exceed the link capacity since the receiver cannot receive packets faster than the link capacity allows. 
}
of both subflows. If there is no fair queuing, the subflow which sends more also achieves a higher throughput (\autoref{fig:illustration_no_fq}), which can be considered unfair. 
In this case, Tonopah doesn't behave different from a network flow which doesn't have fair queuing detection.

@@

The dominant subflow receives $\frac{6}{10}$ of the total allowed sending rate, while the non-dominant one receives $\frac{4}{10}$. 
The dominant subflow and the non-dominant subflow alternate every $\max{(50, \min{(100, \textit{rtt})})}$ milliseconds. 
This time period is also called an \textbf{\textit{interval}}. To determine the presence of fair queuing, the data of the last four intervals are evaluated. 
We look at more than one interval because this reduces the effect of measurement noise. We found four intervals to be a suitable number.

@@

Another safeguard against measurement noise is that we only take a decision regarding fair queuing if the dominant flow actually really sent more data than the non-dominant one. 
Specifically, if the dominant flow sends less than 57.5\%, we don't take a decision because it means that probably we didn't have enough data to send. 
This case only occurs when the underlying application has no data to send or when the sending rate is very low ($<$ 10\,Mbit/s).

@@

Tonopah was implemented on top of Picoquic in C. It was added on top of the NewReno congestion control algorithm but it could easily be added on top of other 
congestion control algorithms as well. By basing our implementation on Picoquic, Tonopah also supports Explicit Congestion Notification.

@@

In the following evaluations we always run each network flow for 90\,s. Tonopah measures whether there is fair queuing or not continuously. 
If in the evaluation scenario there is fair queuing and Tonopah detects fair queuing for 80\,s and fails to detect it for 10\,s, we 
state the accuracy as $\frac{80}{90} = 89\%$. 
\begin{table}
\begin{tabularx}{\columnwidth}{| l | X | X | X |}
\hline
& 10 Mbit/s & 50 Mbit/s & 100 Mbit/s \\ \hline
10ms & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\%\\ \hline
50ms & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% \\ \hline
100ms & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% & Median: 100\% \newline 1st quart.: 100\% \newline 3rd quart.: 100\% \\ \hline
\end{tabularx}
\caption{Detection accuracy in case there's \textbf{no fair queuing}.  The overall median accuracy is 100\%, the first quartile is 100\% and the third quartile is 100\%.}
\label{table:no_fq}
\end{table}

@@

\subsection{Detection accuracy}
